---
title: "Approach 3 - customer profiling"
output: html_document
---

```{r}
# Setup work environments
library(dplyr)
library(data.table)
library(tidyr)
library(ggplot2)
library(cluster)
library(clustMixType)
```

```{r}
# required functions
man_mode <- function(x) {
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}
```

# Customer Clustering
```{r}
# To segment the customers, filter the data for the transactions with Customer ID.
cp$Date_time <- as.POSIXct(cp$Date_time, "%Y-%m-%d %H:%M:%S", tz=Sys.timezone())

cust_attr <- cp[!is.na(cp$Customer.ID),]
```

The feature extraction for the clustering analysis is done as below: 
Extracting hour of the day, day of the week, month and time of the day (Morning, Afternoon & Evening) from the date-time field to understand the seasonality across various levels
```{r}
cust_attr <- cust_attr %>% mutate(month = format(Date_time, '%b'),
                                  day = format(Date_time, '%a'),
                                  hour = format(Date_time, '%H'),
                                  time_of_day = if_else(hour<12, 'Morning',
                                                        if_else(hour<16, 'Afternoon', 'Evening')))
```

Extracting features related to the customers’ purchase behavior such as
o AveragebasketsizeandUniqueitems–tounderstandwhetherthecustomers
buy a variety of products or just loyal to a single product
o Number of visits, Number of weeks and Number of unique weeks – to determine the loyal customers
o MindateandMaxdate–toestimatethetotaltenurethecustomerbeenvisiting o Totalsales,Totalquantity,SalesperweekandQuantityperweek–toidentify
the customers that bring more value
o Number of visits per week – to identify the frequent visitors
o Most purchased category and item – to identify the popular products for various segments of customers
o Usual time of day, Usual day of week and Usual month- to determine the peak hours, days and months
o Total categories and items – to estimate the potential customers for various products
```{r}
cust_attr_1 <- cust_attr%>%group_by(Customer.ID, Date_time)%>%
  summarise(basket_size = sum(Qty),
            sum_sales = sum(Gross.Sales))%>%
  ungroup()%>%
  group_by(Customer.ID)%>%
  summarise(avg_basket = mean(basket_size),
            num_visits = length(Date_time),
            min_date = min(Date_time),
            max_date = max(Date_time),
            total_sales = sum(sum_sales),
            total_qty = sum(basket_size))%>%
  mutate(num_weeks =  if_else(num_visits == 1,1,as.numeric(difftime(max_date,min_date,units='weeks'))),
         num_weeks = if_else(num_weeks < 1, 1, num_weeks),
         sales_per_week = total_sales/num_weeks,
         qty_per_week = total_qty/num_weeks,
         num_visits_per_week = num_visits/num_weeks)
cust_attr_1 <- cust_attr_1[, -c(4,5,7)]
```

```{r}
cust_attr_2 <- cust_attr %>% group_by(Customer.ID) %>% summarise(most_pur_cat = man_mode(Category),
                                                               most_pur_item = man_mode(Item),
                                                               unique_items = n_distinct(Item),
                                                               usual_time = man_mode(time_of_day),
                                                               usual_dow = man_mode(day),
                                                               usual_month = man_mode(month),
                                                               total_qty = sum(Qty),
                                                               total_cat = n_distinct(Category),
                                                               total_item = n_distinct(Item))
```

```{r}
cust_attr_3 <- unique(cust_attr[,c('Customer.ID','Date_time')])%>%
  mutate(year = format(as.Date(Date_time),"%Y"),
         week_num = strftime(as.POSIXct(Date_time),format="%W"),
         year_week = paste0(year,week_num))%>%
  group_by(Customer.ID)%>%
  summarize(num_unique_weeks = n_distinct(year_week))

merge_1 <- merge(cust_attr_1, cust_attr_2, by='Customer.ID')
cust_attr_final <- merge(merge_1, cust_attr_3, by='Customer.ID')

cust_attr_clust <- cust_attr_final[,-1]
```

Now that we have the attributes of customers, we need to normalize the numeric values and discretize the categorical values to bring them all on the same scale. This makes sure that the attributes are given equal importance when the clusters are being formed using them.
```{r}
normalize <- function(x){return ((x - min(x))/(max(x) - min(x)))}

cust_attr_clust <- cust_attr_clust%>%mutate(avg_basket = normalize(avg_basket),
                                            num_visits = normalize(num_visits),
                                            total_sales = normalize(total_sales),
                                            num_weeks = normalize(num_weeks),
                                            sales_per_week = normalize(sales_per_week),
                                            qty_per_week = normalize(qty_per_week),
                                            num_visits_per_week = normalize(num_visits_per_week),
                                            unique_items = normalize(unique_items),
                                            total_qty = normalize(total_qty),
                                            total_cat = normalize(total_cat),
                                            total_item = normalize(total_item),
                                            num_unique_weeks = normalize(num_unique_weeks),
                                            most_pur_cat = as.factor(most_pur_cat),
                                            most_pur_item = as.factor(most_pur_item),
                                            usual_time = as.factor(usual_time),
                                            usual_dow = as.factor(usual_dow),
                                            usual_month = as.factor(usual_month))
```

As we have both categorical and numerical attributes, it is recommended to use k-Prototypes clustering technique. The clusters are formed with the values of k ranging from 1 to 10 and the corresponding SSE is calculated to plot the elbow curve.
```{r}
set.seed(13723) 
SSE_curve <- c() 
for (k in 1:10){
kpro <- kproto(as.data.frame(cust_attr_clust), k, nstarts = 10000)
sse <- sum(kpro$withinss)
  SSE_curve[k] <- sse
}

plot(1:10, SSE_curve, type="b", xlab="Number of Clusters", ylab="SSE")
```

From the above graph, we can see that there is a significant decrease in the SSE of clusters till the value of k is 4. Hence, it would be appropriate to create 4 clusters from the data.
```{r}
set.seed(13723)
kpro <- kproto(as.data.frame(cust_attr_clust), 4, nstarts = 10000)

cust_attr_final$cluster <- kpro$cluster

write.csv(cust_attr_final, "customer_clusters_4.csv", row.names = F)

```

# Customer Profiling based on new clusters
Now that we have clusters formed, we need to understand what kind of customers each cluster have. This will let us understand how to treat or target each of these clusters appropriately. As a first step, we looked at the number of customers in each cluster and their engagement on a weekly basis.
```{r}
library(ggplot2)
ggplot(cust_attr_final, aes(x = cluster, y = n_distinct(Customer.ID)))+
geom_bar(stat = 'identity', color = 'steelblue')+labs(y = 'Number of customers')
ggplot(cust_attr_final, aes(x = cluster, y = num_weeks))+ stat_summary(fun.y = 'mean', geom= 'bar', fill = 'steelblue')
ggplot(cust_attr_final, aes(x = cluster, y = num_visits))+ stat_summary(fun.y = 'mean', geom = 'bar', fill = 'steelblue')
ggplot(cust_attr_final, aes(x = cluster, y = num_visits_per_week))+ stat_summary(fun.y = 'mean', geom= 'bar', fill = 'steelblue')
```

We can see that cluster 3 has the least number of customers but the most loyal ones. Their average number of visits and the average number of weeks in the system is so high when compared with other customers. However, their average number of visits per week is lesser. This means that cluster 3 comprises of loyal customers who have been visiting for a long period of time but at a low frequency.
We should understand how much customers belonging to each of these clusters contribute to the overall revenue and sales quantity.
```{r}
ggplot(cust_attr_final, aes(x = cluster, y = total_sales))+ stat_summary(fun.y = 'mean', geom= 'bar', fill = 'steelblue')
ggplot(cust_attr_final, aes(x = cluster, y = sales_per_week))+ stat_summary(fun.y = 'mean', geom= 'bar', fill = 'steelblue')
ggplot(cust_attr_final, aes(x = cluster, y = total_qty))+ stat_summary(fun.y = 'mean', geom= 'bar', fill = 'steelblue')
ggplot(cust_attr_final, aes(x = cluster, y = qty_per_week))+ stat_summary(fun.y = 'mean', geom= 'bar', fill = 'steelblue')
```

One interesting thing is that even though clusters other than 3 are new customers, cluster 4 seems to be more potential than that of the other clusters. This leaves clusters 3 and 4 to be a priority and need to be targeted according to their interests.
As there is a difference in the behavior of new customers in cluster 4 when compared to those of clusters 1 and 2, we need to further analyze to understand them better. Hence, we looked at the attributes that can uncover the potential of the customers i.e., average basket size and number of unique items bought by customers in this cluster.
```{r}
ggplot(cust_attr_final, aes(x = cluster, y = avg_basket))+ stat_summary(fun.y = 'mean', geom= 'bar', fill = 'steelblue')
```

We establish that the cluster 3 has the most loyal customers and cluster 4 has the potential new customers who can be retained to turn into loyal customers if targeted properly. We expect the behavior of cluster 4 to be like cluster 3. Hence, we looked at the pattern of number of customers visiting the store on a given day and across months.
```{r}
month_trend <- cust_attr_final%>%group_by(cluster, usual_month)%>%summarize(cust_cnt = n_distinct(Customer.ID)) ggplot(month_trend, aes(x = usual_month, y = cust_cnt, group = cluster))+
geom_line( color = 'steelblue')+ ylab("Number of distinct customers")+ xlab("Month")+
facet_wrap(~cluster, scales = "free_x")
```

```{r}
day_trend <- cust_attr_final%>%
  group_by(cluster, usual_time)%>%
  summarize(cust_cnt = n_distinct(Customer.ID)) 

ggplot(day_trend, aes(x = usual_time, y = cust_cnt, group = cluster))+
geom_bar(stat = 'identity', fill = 'steelblue')+ylab("Number of distinct customers")+ xlab("Time of day")+ 
  facet_wrap(~cluster, scales = "free_x")
```
We could see that cluster 3 and 4 are behaving similarly when compared to cluster 1 and 2.

Furthermore, we need to understand the product portfolio of these customers to be able to provide targeted recommendations. As we have seen already, many of the customers in general buy coffee most of the times. Hence, it is better to look at the most purchased products by customers at the cluster level. This will help us understand the purchase preferences of different segments of the customers.

```{r}
cat<- cust_attr_final%>%group_by(cluster, most_pur_cat)%>%summarize(cust_cnt = n_distinct(Customer.ID))
ggplot(cat, aes(x = reorder(most_pur_cat, -cust_cnt), y = cust_cnt, group = cluster))+
geom_bar(stat = 'identity', fill = 'steelblue')+ ylab("Number of distinct customers")+ xlab("most purchased category")+ facet_wrap(~cluster, scales = "free_x")
```
From the above figures, we illustrate that the clusters 2 and 4 behave similarly when compared to other clusters. Customers belonging to these clusters purchase products that belong to other categories as well whereas customers from cluster 1 and 3 purchase coffee most of the times. This gives us an opportunity to up sell and cross sell food items along with coffee, which is the main purchased item in the entire menu.

# Association Rule Mining
We performed association rule mining on individual clusters to identify the products that are most often purchased together. This gives us a sense co-occurrence that exists in the selling of products.
```{r}
library(tidyr) 
library(dplyr)
library(stringr) 
library(arules)
data_total <- read.csv("cleaned_data.csv")
cust_attr_final<- read.csv("customer_clusters_4.csv")
data_ar <- merge(data_total, cust_attr_final[,c("Customer.ID","cluster")], on = "Customer.ID", how = "left")
data_ar <- data_ar[data_ar$cluster==2,]
```

```{r}
cust_level <- data_ar%>%filter(!is.na(Customer.ID))%>% select(Customer.ID,Date_time,Category, Item, Price.Point.Name)%>% group_by(Customer.ID, Date_time)

cust_cat_trans <- unique(cust_level[,c('Customer.ID','Date_time','Item')])

if(sessionInfo()['basePkgs']=="dplyr" | sessionInfo()['otherPkgs']=="dplyr"){ detach(package:dplyr, unload=TRUE)
}
```

```{r}
library(plyr)
cust_cat_trans <- ddply(cust_cat_trans,c("Customer.ID","Date_time"), function(df1)paste(df1$Item,
                                        collapse = ","))
cust_cat_trans$Customer.ID <- NULL 
cust_cat_trans$Date_time <- NULL 
colnames(cust_cat_trans) <- c("itemList")

write.csv(cust_cat_trans,"ItemList.csv", quote = FALSE, row.names = TRUE)
```

```{r}
txn = read.transactions(file="ItemList.csv", rm.duplicates= TRUE, format="basket",sep=",",cols=1);
txn@itemInfo$labels <- gsub("\"","",txn@itemInfo$labels)
```

```{r}
rules <- apriori(txn,parameter = list(sup = 0.01, conf = 0.1)) 
rules <- sort(rules, by = "lift", decreasing = T)
inspect(rules)
```

